Dependent Types for Safe and Secure Web Programming

### 2014-02-03 paper/conclusion.tex
Few more revisions

Java is often used as a language to write enterprise-level web applications. Frameworks such as the Java Persistence API (JPA) \cite{jpa} use Java annotations to translate data models into the appropriate database schemata, but such code may soon become unwieldy due to the large amount of redundant boilerplate code such as accessor and mutator functions. WebDSL \cite{webdsl} is a domain-specific language written primarily to
introduce new abstractions which aim to reduce the amount of boilerplate code
that must be written and maintained by developers. 
The DSL is parsed into an abstract syntax tree, modified using rewrite rules, and elaborated back into Java code. WebDSL also applies similar concepts to implement a \textit{template system} for the presentation of data entities. We look to
implement many of these ideas, but as effects within the IdrisWeb
framework, as with the form construction effect.

@@

Our framework provides several static
guarantees. Data submitted by users is inherently unsafe and
means systems are vulnerable to attacks
such as SQL injection. This particular threat is ameliorated due
to elements being associated with specific types during form construction. This
immediately eliminates the possibilities of SQL injection attacks on non-string
types. Since failures are handled transparently, no runtime errors are output
to the browser, meaning that attackers may not use such information to aid
attacks. Additionally, since checking is performed on the types of the form
elements and the types of arguments accepted by the handler, it is impossible
to associate a form with a handler incompatible with the submitted data.

@@

We have shown that embedded domain-specific languages using dependent types and
algebraic effects can be used to increase confidence in web applications by
providing additional static guarantees about runtime behaviour, but much more
can be done using the same approach.

@@

Since at this stage we have concentrated on the use of dependent types for
enforcing resource usage protocols and type-safe form handling, we currently
handle the generation of HTML in an unstructured manner. Future work will
entail a DOM library to facilitate the generation and manipulation of HTML, in
turn giving stronger guarantees about its correctness. Other planned features
include a template system, allowing for web pages to be automatically generated
from data, and an object-relational mapping system which will allow users to
manipulate records which can then be automatically written to the database,
instead of having to update tables manually via SQL queries.

>>>>

Ensuring conformance to resource usage protocols has been attempted using the notion of \emph{typestate} in object-oriented languages \cite{deline:typestates}. Our approach differs in that we make no changes to the type system in order to implement this functionality: we check resource usage protocol conformance purely within \idris{}.
Additionally, our handler-based approach facilitates greater control over side effects.

Java is often used as a language to write enterprise-level web applications. Frameworks such as the Java Persistence API (JPA) \cite{jpa} use Java annotations to translate data models into the appropriate database schemata, but such code may soon become unwieldy due to the large amount of redundant boilerplate code such as accessor and mutator functions. WebDSL \cite{webdsl} is a domain-specific language written primarily to
introduce new abstractions which aim to reduce the amount of boilerplate code
that must be written and maintained by developers. 
The DSL is parsed into an abstract syntax tree, modified using rewrite rules, and elaborated back into Java code. 
%WebDSL also applies similar concepts to implement a \textit{template system} for the presentation of data entities. We look to
%implement many of these ideas, but as effects within the IdrisWeb
%framework, as with the form construction effect.

@@

Our framework provides several static
guarantees. Data submitted by users is inherently unsafe and
means systems are vulnerable to attacks
such as SQL injection. This particular threat is ameliorated due
to elements being associated with specific types during form construction. This
immediately eliminates the possibilities of SQL injection attacks on non-string
types. Since failures are handled transparently, no runtime errors are written
to the browser, meaning that attackers may not use such information to aid
attacks. Additionally, since checking is performed on the types of the form
elements and the types of arguments accepted by the handler, it is impossible
to associate a form with a handler incompatible with the submitted data.

@@

We have shown that embedded domain-specific languages using dependent types and
algebraic effects can be used to increase confidence in web applications, but much more
can be done using the same approach.

@@

Since at this stage we have concentrated on the use of dependent types for
enforcing resource usage protocols and type-safe form handling, we currently
handle the generation of HTML in an unstructured manner. Future work will
entail a DOM library to facilitate the generation and manipulation of HTML, in
turn giving stronger guarantees about its correctness. Other planned features
include a template system, allowing for web pages to be automatically generated
from data, and an object-relational mapping system. 
It would also be interesting to 
explore the use of continuation-passing approaches \cite{queinnec:ioc} to enhance the usability of the framework.

### 2014-02-03 paper/protocols.tex
Few more revisions

%Programs making use of the DSL should look familiar to developers even without
%a background in functional programming. 
To demonstrate the 
library, we return to the previous example of selecting the names and addresses
of all staff working in a given department. 
%Since the
%\texttt{Effects} library overloads the bind operator, we may make use of
%\texttt{do}-notation, facilitating the usage of an imperative style.
We define a function \texttt{textSel} of type:

@@

\begin{SaveVerbatim}{testsel}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either QueryError (List (String, String)))
testSelect dept = do
  db_res <- openDB "people.db"
  if_valid then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if_valid then do 
      bindText 1 dept
      bind_res <- finishBind
      if_valid then do
        executeStatement
        results <- collectResults
        finaliseInvalid
        closeDB
        Effects.pure $ Right results
      else do
        cleanupBindFail
        Effects.pure $ Left (getBindError bind_res)
    else do
      cleanupPSFail
      Left $ getQueryError sql_prep_res
  else do 
    Effects.pure $ Left (getQueryError db_res)
\end{SaveVerbatim}

>>>>

%Programs making use of the DSL should look familiar to developers even without
%a background in functional programming. 
To demonstrate the 
library, we return to the previous example of selecting the names and addresses
of all staff working in a given department. 
%Since the
%\texttt{Effects} library overloads the bind operator, we may make use of
%\texttt{do}-notation, facilitating the usage of an imperative style.
We define a function \texttt{textSelect} of type:

@@

\begin{SaveVerbatim}{testsel}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either QueryError (List (String, String)))
testSelect dept = do
  db_res <- openDB "people.db"
  if_valid then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if_valid then do 
      bindText 1 dept
      bind_res <- finishBind
      if_valid then do
        executeStatement
        results <- collectResults
        finaliseInvalid
        closeDB
        return $ Right results
      else do
        cleanupBindFail
        return $ Left (getBindError bind_res)
    else do
      cleanupPSFail
      return $ Left (getQueryError sql_prep_res)
  else do 
    return $ Left (getQueryError db_res)
\end{SaveVerbatim}
